Simulating the Linux page cache in distributed applications with WRENCH

### 2020-12-09 paper_io_sim.tex

\title{Simulating the Linux page cache in distributed applications with the WRENCH library}

@@

        The rapid emergence of Big Data seen in recent years has resulted in a 
        growing need for efficient data storing and optimized data analysis techniques. 
        Despite relying on large computing infrastructures with powerful computing capability 
        to process these large datasets, an I/O bottleneck remains. 
        One widely-used approach to minimize the costs of data transfers during
        regular use is through the employment of the page cache.
        While many simulation tools have been developed to 
        support studies in computing systems optimization, page cache simulation 
        has not been fully captured or remains absent from these simulators. 
        The inability to simulate I/O with page cache leads to inaccurate 
        simulations in many situations, especially when dealing with data intensive applications.

        In this paper, we propose a I/O simulation model that characterizes the key features of the page cache.
        We have implemented the model with SimGrid, a simulation framework for 
        distributed systems, using WRENCH, a SimGrid-based workflow simulator. 
        Our simulator is compatible with both single-threaded and multithreaded applications,
        and can simulate both writeback and writethrough cache that can factor into local
        and network-based filesystems.

@@

        Characterizing the performance of applications directly on computing infrastructure
        has various shortcomings, which include, limited and shared resources, operational
        costs, unstable environments, and the difficulties in producing reproducible results.
        Simulations address these concerns by providing abstractions for the performance of computer
        hardware, such as CPU, network and storage. Consequently, simulations provide and
        cost-effective, fast, easy and reproducible  way evaluate application performance
        on different infrastructures without having to execute the code.

@@

        Page cache simulation is crucial when simulating data intensive applications,
        which rely on cache hits and writing to memory for additional performance.
        Furthermore, simulating the page cache enables us to study important application-I/O
        bottlenecks which can arise over the course of execution, such as page cache
        starvation~\cite{zhuang2017}.
        While existing simulators can adequately characterize many different computing environments,
        they lack much of the ability to adequately describe the page cache. Some
        simulators try to simulator I/O with page cache, but fail to capture the key features such
        as dirty data and cache eviction policies~\cite{nunez2012simcan,nunez2012icancloud}. 
        Other simulators, such as~\cite{xu2018}, capture the important feature of the
        page cache, but are domain-specific.

@@

        In this paper, we present WRENCH-Cache, a simulation model of I/O with page
        cache incorporated into WRENCH~\cite{casanova2018wrench}, a workflow management system simulator based on
        the popular SimGrid framework. The goal of WRENCH-Cache is to provide users
        with an accurate representation of a workflow execution, when file system page
        cache is used, while providing a good balance between accuracy and scalability.
        WRENCH-Cache is able to simulate both writeback and writethrough
        cache which would normally be available to local and NFS-based filesystems.
        WRENCH-Cache can provide accurate simulations for large-scale parallel
        applications using the page cache.

@@

        To improve performance of disk accesses, the page cache was introduced to the Linux kernel. %It contains pages referring to physical pages on disk \cite{linuxdev3rd2010}.
            The goal of the page cache is to try ensure that most I/O operations occur entirely
            within memory space rather than synchronously to and from the storage device.
            All disk-backed data located in the page cache are partitioned into many small pages.
            When a read operation is initiated, the kernel checks whether the required pages are in memory.
            If it is the case, data is read entirely from memory (i.e. \textit{cache hit}). Otherwise,
            the necessary pages must be loaded to memory from the disk (i.e. \textit{cache miss})
            Conversely, for writing, one of three strategies can be applied: direct~I/O (or, ``no-write''),
            writethrough and writeback.
            If direct~I/O is enabled for a given file system, all I/O operations bypass the
            page cache for writes, writing data directly to disk. In contrast, write-through enables the page cache
            for file system writes, but the data is written synchronously from memory to disk. Writeback extends write-through
            by enabling asynchronous writes to disk, meaning that a write call will return once
            the data has been written to memory.
            The writeback strategy is considered to outperform write-through as well as 
            direct I/O as it delays disk writes to perform a bulk write at a later time.
            \cite{linuxdev3rd2010}.

@@

            Many simulation frameworks exist to describe the behaviours of applications
            running on the various infrastructures available to users, such as personal computers,
            clouds and clusters.
            These simulators are built with resource models such CPU, network and 
            storage, using various different approaches to describe the hardware.
            For example, there is \textit{CloudSim} and \textit{iCanCloud} for packet-level network models 
            and \textit{DiskSim} the simulates storage hardware functioning.

            Simulators must choose between scalability and accuracy when describing the
            hardware.\cite{casanova2014simgrid}. 
            Fine-grained models allow simulators to achieve high accuracy, 
            but lead to scalability issues due to long simulation time 
            when simulated platforms and applications grow. 
            SimGrid is an open source simulation platform that has been developed 
            not only to balance the trade-off between accuracy and scalability, 
            but also to achieve versatility, allowing the use of the framework 
            to simulate applications in multiple domains. WRENCH, a workflow management
            system simulator based on the SimGrid framework, provides a high-level API
            for simulating workflows, thus facilitating the integration of new models.
            \valerie{say more about wrench}

@@

    \section{Method}
    \label{method}
    In this section, we present our page cache simulation model for file I/O,
    flusing, and eviction. We implement this model in two simulators:
    (1) a standalone Python prototype, and (2) WRENCH-cache, an extension of the WRENCH library.
    We also describe experimental scenarios to evaluate our simulators in various configurations.

    We separate our simulation model in two components, the IO
    Controller and the Memory Manager, which together simulate 
    file reads and writes in Linux (Figure~\ref{fig:interaction}). 
    To read or write a file chunk, a simulated application sends a
    request to the IO Controller. The IO Controller interacts as needed with
    the Memory Manager to free memory through flushing or eviction,
    and to read or write to cache. The Memory Manager
    implements these operations, simulates periodical flushing
    and eviction from cache, and reads or writes to disk when necessary.  
    In case write-through strategy is used, IO Controller can also interact directly 
    with disk model to simulate disk writes.

@@

    The Memory Manager simulates two parallel threads: the main one
    implements flushing, eviction, and cache I/Os synchronously, whereas
    the second one, operating in the background, periodically searches for
    expired dirty data in page cache LRU lists and flushes them to disk. We
    use existing storage simulation models~\cite{lebre2015} to simulate disk and
    memory, characterized by its storage capacity, read and write
    bandwidths, and latency. Reusing existing storage models allows us to
    simulate bandwidth sharing between concurrent memory accesses and concurrent disk accesses.

@@

        \tristan{discussion could be fleshed out}
\bibliographystyle{plain}
\bibliography{citation}

>>>>

\title{Simulating the Linux page cache in distributed applications with WRENCH}

@@

        The rapid emergence of Big Data seen in recent years has resulted in a 
        growing need for efficient data storing and optimized data analysis techniques. 
        Despite relying on large computing infrastructures with powerful computing capability 
        to process these large datasets, an I/O bottleneck remains. 
        One widely-used approach to minimize the costs of data transfers during
        regular use is through the employment of the Linux page cache.
        While many simulation tools have been developed to 
        support studies in computing systems optimization, page cache simulation 
        has not been fully captured or remains absent from these simulators. 
        The inability to simulate I/O with page cache leads to inaccurate 
        simulations in many situations, especially when dealing with data-intensive applications.

        In this paper, we propose a I/O simulation model that characterizes
        the key features of the page cache. We have implemented the model
        with SimGrid, a simulation framework for distributed systems, using
        WRENCH, a SimGrid-based workflow simulator. Our simulator is
        compatible with both single-threaded and multithreaded
        applications, and can simulate both writeback and writethrough
        cache that can factor into local and network-based filesystems. We
        evaluate it in different conditions, including sequential and
        concurrent applications, as well as local and remote I/Os.
        The page cache model reduces the simulation
        error by up to an order of magnitude compared to cacheless simulations.

@@

        Characterizing the performance of applications directly on computing infrastructure
        has various shortcomings, which include, limited and shared resources, operational
        costs, unstable environments, and the difficulties in producing reproducible results.
        Simulations address these concerns by providing abstractions for the performance of computer
        hardware, such as CPU, network and storage. Consequently, simulations provide a
        cost-effective, fast, easy and reproducible  way evaluate application performance
        on different infrastructures without having to execute the code.

@@

        Page cache simulation is crucial when simulating data intensive applications,
        which rely on cache hits and writing to memory for additional performance.
        Furthermore, simulating the page cache enables us to study important application-I/O
        bottlenecks which can arise over the course of execution, such as page cache
        starvation~\cite{zhuang2017}.
        While existing simulators can adequately characterize many different computing environments,
        they lack much of the ability to adequately describe the page cache. Some
        simulators try to simulate I/O with page cache, but fail to capture the key features such
        as dirty data and cache eviction policies~\cite{nunez2012simcan,nunez2012icancloud}. 
        Other simulators, such as~\cite{xu2018saving}, capture the important features of the
        page cache, but are domain-specific.

@@

        In this paper, we present WRENCH-cache, a simulation model of I/O with page
        cache incorporated into WRENCH~\cite{casanova2018wrench}, a workflow management system simulator based on
        the popular SimGrid framework. The goal of WRENCH-cache is to provide users
        with an accurate representation of a workflow execution, when file system page
        cache is used, while providing a good balance between accuracy and scalability.
        WRENCH-cache is able to simulate both writeback and writethrough
        cache which would normally be available to local and NFS-based filesystems.
        WRENCH-cache can provide accurate simulations for large-scale parallel
        applications using the page cache.

@@

        To improve performance of disk accesses, the page cache was introduced to the Linux kernel. %It contains pages referring to physical pages on disk \cite{linuxdev3rd2010}.
            The goal of the page cache is to try ensure that most I/O operations occur entirely
            within memory space rather than synchronously to and from the storage device.
            All disk-backed data located in the page cache are partitioned into many small pages.
            When a read operation is initiated, the kernel checks whether the required pages are in memory.
            If it is the case, data is read entirely from memory (i.e. \textit{cache hit}). Otherwise,
            the necessary pages must be loaded to memory from the disk (i.e. \textit{cache miss})
            Conversely, for writing, one of three strategies can be applied: direct~I/O (or, ``no-write''),
            writethrough and writeback.
            If direct~I/O is enabled for a given file system, all I/O operations bypass the
            page cache for writes, writing data directly to disk. In contrast, write-through enables the page cache
            for file system writes, but the data is written synchronously from memory to disk. Writeback extends write-through
            by enabling asynchronous writes to disk, meaning that a write call will return once
            the data has been written to memory.
            The writeback strategy is considered to outperform write-through as well as 
            direct I/O as it delays disk writes to perform a bulk write at a later time
            \cite{linuxdev3rd2010}.

@@

            Many simulation frameworks exist to describe the behaviour of applications
            running on the various infrastructures available to users, such as personal computers,
            clouds and clusters.
            These simulators are built with resource models such CPU, network and 
            storage, using various different approaches to describe the hardware.
            For example, there is \textit{CloudSim} and \textit{iCanCloud} for packet-level network models 
            and \textit{DiskSim} the simulates storage hardware functioning \tristan{add references}.

            Simulators must choose between scalability and accuracy when describing the
            hardware~\cite{casanova2014simgrid}. 
            Fine-grained models allow simulators to achieve high accuracy, 
            but lead to scalability issues due to long simulation time 
            when simulated platforms and applications grow. 
            SimGrid is an open source simulation platform that has been developed 
            not only to balance the trade-off between accuracy and scalability, 
            but also to achieve versatility, allowing the use of the framework 
            to simulate applications in multiple domains. WRENCH, a workflow management
            system simulator based on the SimGrid framework, provides a high-level API
            for simulating workflows, thus facilitating the integration of new models.
            \valerie{say more about wrench}

@@

    \section{Method}
    \label{method}
    We separate our simulation model in two components, the IO
    Controller and the Memory Manager, which together simulate 
    file reads and writes (Figure~\ref{fig:interaction}). 
    To read or write a file chunk, a simulated application sends a
    request to the IO Controller. The IO Controller interacts as needed with
    the Memory Manager to free memory by triggering flushing or eviction,
    and interact with cached data. The Memory Manager
    implements these operations, simulates periodical flushing
    and eviction from cache, and reads or writes to disk when necessary.
    In case the write-through strategy is used, the IO Controller can also interact directly 
    with disk model to simulate disk writes.

@@

    The Memory Manager simulates two parallel threads: the main one
    implements flushing, eviction, and cached I/Os synchronously, whereas
    the second one, operating in the background, periodically searches for
    expired dirty data in page cache LRU lists and flushes them to disk. We
    use existing storage simulation models~\cite{lebre2015} to simulate disk and
    memory, characterized by its storage capacity, read and write
    bandwidths, and latency. Reusing existing storage models allows us to
    simulate bandwidth sharing between concurrent memory accesses and concurrent disk accesses.

@@

        \tristan{discussion could be fleshed out}

        \section{Acknowledgments}
The computing platform used in the experiments was obtained with funding
from the Canada Foundation for Innovation.
\bibliographystyle{plain}
\tristan{Check references, in particular capitalization}
\bibliography{citation}
